MSstatsPTM: Statistical relative quantification of post-translational modifications in bottom-up mass spectrometry-based proteomics

### 2022-06-16 main/MSstatsPTM_Main.tex
fixed some typos

\todo{DK: I'm not sure where this part should be. could move it to the first paragraph of this section, or in the extensions to complex design}
Finally, this approach allows us to model experiments with separate designs for the PTM and global profiling runs. This includes differing number of runs, biological replicates, or acquisition methods. This is achieved through fully modeling the PTM and global protein separately before adjusting. Because we are not combining on the replicate level, the number of replicates for each group do not need to match for adjustment. This makes the proposed method more flexible and applicable to a wider variety of experimental designs.

@@

The proposed methods are implemented in the open source R package $MSstatsPTM$, available on Bioconductor and Github. An overview of the steps of the package are illustrated in Figure \ref{fig:msstatsptm_design}. $MSstatsPTM$ includes converters for multiple spectral processing tools, including MaxQuant, Progenesis, and Spectronaut. The converters take as input the raw data from the tool, identify the modification site for modified peptide, and put the data into the correct format for analysis in $MSstatsPTM$. Conversion is done separately for the modified peptide and global profiling runs. If the global profiling information is not available, the package can still analyze the modified run, but will do so without adjusting for changes in unmodified protein abundance. Specifically, this will reduce our null hypotheses to Equation \ref{eq:conv_null_hyp}; which amounts to running the methods seen in $MSstats$ on modified peptide level data \cite{Choi:2014} \cite{Huang:2020}.

After using the converters, the next step is peptide/protein summarization and missing value imputation. For the modified run, the package summarizes features which include the same modification together. Features with multiple modification sites are not included with single site features and are summarized separately. For the global profiling run all unmodified features from the same protein are summarized together, up to the protein level. Additionally, the summarization includes global median normalization and normalizes between MS runs. The package can optionally impute missing values using an accelerated failure time (AFT) model~\cite{Wei:1992}. Missing value imputation should be performed with care when applied to PTM data. PTMs generally exhibit low feature counts and may be missing due to reasons other than low abundance. These issues can cause the missing value imputation to be unstable.

@@

Beyond the core functionality, the package also includes functions for plotting the results. These include summarization plots to help with quality assurance and identifying sources of variation (such as quality control plots and profile plots), as well as modeling plots (such as volcano plots and heatmaps), which evaluates the fold change between conditions and peptides.

>>>>

\todo{DK: I'm not sure where this part should be. could move it to the first paragraph of this section, or in the extensions to complex design}
Finally, this approach allows us to model experiments with separate designs for the PTM and global profiling runs. This includes differing number of runs, biological replicates, or types of acquisition methods. This is achieved through fully modeling the PTM and global protein separately before adjusting. Because we are not combining on the replicate level, the number of replicates for each group do not need to match for adjustment. This makes the proposed method more flexible and applicable to a wider variety of experimental designs.

@@

The proposed methods are implemented in the open source R package $MSstatsPTM$. An overview of the steps of the package are illustrated in Figure \ref{fig:msstatsptm_design}. $MSstatsPTM$ includes converters for multiple spectral processing tools, including MaxQuant, Progenesis, and Spectronaut. The converters take as input the raw data from the tool, identify the modification site for modified peptide, and put the data into the correct format for analysis in $MSstatsPTM$. Conversion is done separately for the modified peptide and global profiling runs. If the global profiling information is not available, the package can still analyze the modified run, but will do so without adjusting for changes in unmodified protein abundance. This will reduce our null hypotheses to Equation \ref{eq:conv_null_hyp}; which amounts to running the methods seen in $MSstats$ on modified peptide level data \cite{Choi:2014} \cite{Huang:2020}.

After using the converters, the next step is peptide/protein summarization and missing value imputation. For the modified run, the package summarizes features which include the same modification together. Features with multiple modification sites are not included with single site features and are summarized separately. For the global profiling run all unmodified features from the same protein are summarized together, up to the protein level. Additionally, the summarization includes global median normalization and normalizes between MS runs. The package can optionally impute missing values using an AFT model~\cite{Wei:1992}. Missing value imputation should be performed with care when applied to PTM data. PTMs generally exhibit low feature counts and may be missing due to reasons other than low abundance. These issues can cause the missing value imputation to be unstable.

@@

Beyond the core functionality, the package also includes functions for plotting the results. These include summarization plots to help with quality assurance and identifying sources of variation (such as quality control plots and profile plots), as well as modeling plots (such as volcano plots and heatmaps), which evaluate the fold change between conditions and peptides.
